---
title: "Iteration - lots of different methods"
author: "Rosemary Hartman and Jason DuBois"
date: "03/20/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = F)
library(tidyverse)
library(knitr)
#purl("dataset restructuring.Rmd", output="restructure.R")
#source("Data manipulation/restructure.R")
```


##	Loops

We've now got our zooplankton data set in a format we can work with, but now we probably want to do some calculations on it. What if we want to calculate annual average catch?

We can do it one year at a time:
```{r}

mean(filter(CBlong, Year == 1984)$CPUE)

mean(filter(CBlong, Year == 1985)$CPUE)

mean(filter(CBlong, Year == 1986)$CPUE)

```
But we don't want to keep copying and pasting! Instead, we can use a loop. For some items in variable `x`, do something.

### `for` Loop

```
for (items in variable) {
  # do something
}
```

...an example

```{r for-loop}
for (i in 1:5) {
  print(i)
}
```

to calculate the means:

```{r}

#first we set up an empty data frame to accept our output
output = data.frame(year = min(CBlong$Year):max(CBlong$Year), Mean = NA)


for (i in 1:nrow(output)) {   # sequence
  
  output$Mean[i] = mean(filter(CBlong, 
                               Year == output$year[i])$CPUE, na.rm = T)  #body
}

```

That took a few seconds. You may have heard "don't use for loops in R - they are slow!" That is true when you are dealing with very large data sets, but remember: If it works, it's good code! However, you might want to try some different approaches to speed up your data processing, or make your code more readable.


## Apply

The `apply` family of functions includes a number of "wrappers" for loops that make them run faster with less code involved. 
Let's start with a quick look at the help documentation for `apply`

```{r}
# `<_>apply` Family Functions
?apply

```
There are some shortcut fucntions for using `apply` on certain data types. For example, you can use `lapply` for lists and `vapply` for vectors. The basic syntax is:

Here in `lapply` the argument to `FUN` is that 'do something' in the `for` loop. The function passed to `FUN`, will act on every item in the `list or vector`. 

```
lapply(list or vector, FUN = function(x) # do something)
```


### `lapply` First Example

Here we supply the vector 1-->5. We wish to add 5 to each number in that vector.

```{r fun-lapply}

lapply(1:5, FUN = `+`, 5)

```
But with this example, we can do the same thing without using `lapply`. But because R is vectorized, we can simply write below. The results remain neatly in a vector.

```{r vectorized}
1:5 + 5
```

### `lapply` On List

But what if instead of a vector, we had a **list** of vectors? (an uneven list in this case)


```{r list-create}

X <- list(
  Item1 = 1:5,
  Item2 = 6:10,
  Item3 = 11:20
)

X

```

because `X` is a list, vectorization concept doesn't work...

```{r list-add5, eval=FALSE}

X + 5
# Error in X + 5 : non-numeric argument to binary operator

```

so, here's where `lapply` is useful. One nice thing about this is you don't have to set up an empty data frame to save your work the way you do for a loop.


```{r list-lapply}

lapply(X, FUN = `+`, 5)
lapply(X, FUN = `^`, 2)

```


### Now let's try it with vectors using  `vapply`

```{r employees-list}

employees <- list(
  c("first", "last"),
  c("first", "last"),
  c("first", "last"),
  c("first", "last"),
  c("first")
)

employees

```

can we get employees' first name or concatenate `first` & `last`...

```{r employees-demo}

vapply(employees, FUN = `[`, FUN.VALUE = character(1L), 1)
vapply(employees, FUN = paste0, FUN.VALUE = character(1L), collapse = " ")

```

let's display our numeric list again

```{r X-display}

# display list `X` again for convenience
X

```

...what if we wanted some descriptive stats for each item in `X`


```{r vapply-matrix_out1} 

vapply(X, FUN = function(x) {
  c(Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(3L))

```


### List of Functions

We could pass a list of functions as an argument to `vapply`'s `X`. Though this convention is not as common. *Note*: the last function in the list is a custom function that returns count of values less than 5. `x < 5` returns a Boolean, and summing a Boolean returns number `TRUE`, as `TRUE` = 1, `FALSE` = 0.

```{r list-functions}

funs <- list(
  Mean = mean,
  Sd = sd,
  N = length,
  NumValsLT5 = function(x) sum(x < 5)
)

vapply(funs, FUN = function(f, nums) {
  f(nums)
}, FUN.VALUE = numeric(1L), nums = 1:10)

```


# Clean Data

Now let's take a look at some of the zooplankton data! The first thing we want to know about our data is...is it clean? Where are the NA values (in which fields)? What are field data types? Are these the correct data types? ...

`vapply` can help with this.

In this first example, we check each field in `Zoop` for NA values. The function will return 0 if no values are NA, and a value > 0 otherwise. The value > 0 indicates the number of NA records in that field.

```{r na-check1}

vapply(Zoop, FUN = function(x) {
  sum(is.na(x))
}, FUN.VALUE = numeric(1L))

```

Here, we include `typeof` to our output. `vapply` no returns a matrix. *Note*: we use `base::noquote` to remove quotes from the output (`typeof` returns a character), and `base::t` to transpose the matrix as two columns instead of two rows.

```{r na-check2}

t(noquote(vapply(Zoop, FUN = function(x) {
  c(NumNA = sum(is.na(x)), TypeOf = typeof(x))
}, FUN.VALUE = character(2L))))

```

##  Custom Data Summary

Of course, using `utils::str` or `base::summary` on new data is always a good idea. Here, though, we can customize our summary, if you will. For each field in `Zoop`, if value is a double, return the range. Otherwise, return frequency table.


```{r summary-plus}

lapply(Zoop, FUN = function(x) {
  if (is.double(x)) return(range(x, na.rm = TRUE))
  table(x, useNA = "ifany", dnn = NULL)
})

```

We notice something interesting about fields `EZStation` and `DWRStation`. We have some quoted NAs. That is, entries in the Excel file are `NA`, not just blank (which would be read by R as `r NA`). This certainly poses a problem for data summary and other analytics.

```{r na-quoted}

vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L))

```

# Note for Jason: this seems needlessly complicated

So how do we efficiently replace the quoted (or character) NA with R's `r NA`?

(1) `char_nas` returns the number of character NAs in each field
(2) `b` sets to `TRUE` any value > 0; `head(Zoop[b])` displays n=6 for quick viewing
(3) change all character NAs to `r NA` using `lapply`

*Note*: using `Zoop[b]` before assignment `<-` maintains the data.frame structure. Otherwise, you will get a list.

*Note*: instead of anonymous function, below would be much better suited as a named function. This would allow for ease of use on other data.frames

```{r char-nas}

char_nas <- vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L), USE.NAMES = FALSE)

b <- char_nas > 0

head(Zoop[b])

Zoop[b] <- lapply(Zoop[b], FUN = function(x) {
  x[x %in% "NA"] <- NA
  x
})
```

# Here's a simpler way to do it:

The function "na_if" automatically replaces a certain value with an NA

```{r}

Zoop2 = mutate(Zoop, EZStation == na_if(EZStation, "NA"),
               DWRStation == na_if(DWRStation, "NA"))

```


# Split-apply-combine

We often want to split up data sets based on one column and apply functions to another columns. For example, if we wanted to calculate average mysid catch by year, we can use the functions "split" and then "vapply". 

```{r mean-temp-yr1}

mean_temp_yr <- with(data = Zoop, expr = {
  
  s <- split(Temperature, f = Year)
  
  vapply(s, FUN = mean, FUN.VALUE = numeric(1L))
  
})

mean_temp_yr

```

The package dplyr has some nifty tools to do the same thing. In this case, we want to group our data frame based on year and apply the "mean" fuction to CPUE. 

```{r}

CBmeans = CBlong %>%
  group_by(Year) %>% #group it by Year
  summarize(MeanCPUE = mean(CPUE)) #Calculate Mean CPUE

CBmeans
#much faster than the loop!


#we can get really complicated if we want
CBmeans = CBlong %>%
  group_by(Year) %>% #group it by Year
  summarize(MeanCPUE = mean(CPUE), #Calculate Mean CPUE
                    sdCPUE = sd(CPUE), #calculate standard deviation
                    nobs = length(CPUE), #number of observations
                    seCPUE = sdCPUE/nobs) #calculate standard error
View(CBmeans)

#note that you can use your new variables immediately!
```                    


## Now it's your turn

1. What is the average CPUE of each species?

2. What is the maximum CPUE of Eurytemora in Suisun Bay by year?

3. Calculate the relative % composition of each species by year (this is a hard one).
