---
title: ''
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Style Guide

PascalCase: function names; dataframes; field headings; file names  
camelCase: function parameters  
lower_case (with underscore): variables  

use `<-` instead of `=` for variable assignment  
most everytime function call, hard return after left parentheses  
each line keep to max 80 characters  

```
# create function
LengthFrequency <- function(data, speciesName, forkLength) {
  # do something here
}
# end LengthFrequency

len_freq <- LengthFrequency(
  data = Sturgeon,
  speciesName = "white",
  # speciesName = "green",
  forkLength = TRUE
)
```

# Functional Programming

- function call
- arguments can be other functions
- functions can return functions

# Function Call

`iris` built-in dataframe in `datasets` package

```{r func-call}

`+`(9, 8)                      # 9 + 8

`<-`(x, 1:10)                  # x <- 1:10 or x = 1:10
print(x)

`==`(9, 8)                     # 9 == 8

`[`(iris, 1, 2)                # iris[1, 2]
`[[`(iris, 1, 2)               # iris[[1, 2]]

`[[`(iris, 1, "Sepal.Width")   # iris[[1, "Sepal.Width"]]

# Not run --------------------------------------------------
# `[`(iris, 5)                 # iris[5]
# `[`(iris, , 5, drop = TRUE)  # iris[, 5, drop = TRUE]
# `[[`(iris, 5)                # iris[[5]]



```

# `for` Loop

```
for (items in variable) {
  # do something
}
```

```{r for-loop}
for (i in 1:5) {
  print(i)
}
```

# `<_>apply` Family Functions

```
lapply(list or vector, FUN = function(x) # do something)
```

```{r lapply-show}

lapply

```

```{r fun-lapply}

lapply(1:5, FUN = `+`, 5)

```

```{r vectorized}
1:5 + 5
```

```{r list-create}

X <- list(
  Item1 = 1:5,
  Item2 = 6:10,
  Item3 = 11:20
)

```

```{r list-add5, eval=FALSE}

X + 5
# Error in X + 5 : non-numeric argument to binary operator

```

```{r list-lapply}

lapply(X, FUN = `+`, 5)
lapply(X, FUN = `^`, 2)

```

```{r length-X}

lapply(X, FUN = length)

```

```{r vapply-show}

vapply

```

```{r vapply-example}

vapply(X, FUN = length, FUN.VALUE = numeric(1L))

```

```{r employees-list}

employees <- list(
  c("first", "last"),
  c("first", "last"),
  c("first", "last"),
  c("first", "last"),
  c("first")
)

employees

```

```{r employees-demo}

vapply(employees, FUN = `[`, FUN.VALUE = character(1L), 1)
vapply(employees, FUN = paste0, FUN.VALUE = character(1L), collapse = " ")

```

```{r X-display}

# display list `X` again for convenience
X

```

```{r vapply-matrix_out1}

vapply(X, FUN = function(x) {
  c(Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(3L))

```

```{r vapply-matrix_out2}

vapply(X, FUN = function(x) {
  c(N = length(x), Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(4L))

```

```{r vapply-matrix_out3}

t(vapply(X, FUN = function(x) {
  c(N = length(x), Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(4L)))

```

```{r vapply-list_out, eval=FALSE, echo=FALSE}

t(vapply(X, FUN = function(x) {
  list(N = length(x), Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = list(N = NA, Min = NA, Max = NA, Mean = NA)))

```

```{r list-functions}

funs <- list(
  Mean = mean,
  Sd = sd,
  N = length,
  NumValsLT5 = function(x) sum(x < 5)
)

vapply(funs, FUN = function(f, nums) {
  f(nums)
}, FUN.VALUE = numeric(1L), nums = 1:10)

```

```{r data-download}

tf <- tempfile(pattern = "thisone", fileext = ".xlsx")

download.file(
  url = "ftp://ftp.wildlife.ca.gov/IEP_Zooplankton/1972-2018MysidMatrix.xlsx",
  # destfile = tempfile(fileext = ".xlsx")
  destfile = tf,
  mode = "ab"
)

Zoop <- readxl::read_excel(
  path = tf,
  sheet = "Mysid CPUE Matrix 1972-2018 "
)

unlink(tf)

```

```{r na-check}

vapply(Zoop, FUN = function(x) {
  sum(is.na(x))
}, FUN.VALUE = numeric(1L))

t(noquote(vapply(Zoop, FUN = function(x) {
  c(NumNA = sum(is.na(x)), TypeOf = typeof(x))
}, FUN.VALUE = character(2L))))

```

```{r summary-plus}

lapply(Zoop, FUN = function(x) {
  if (is.double(x)) return(range(x, na.rm = TRUE))
  table(x, useNA = "ifany", dnn = NULL)
})

```

```{r na-quoted}

vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L))

```

```{r char-nas}

char_nas <- vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L), USE.NAMES = FALSE)

b <- char_nas > 0

head(Zoop[b])

Zoop[b] <- lapply(Zoop[b], FUN = function(x) {
  x[x %in% "NA"] <- NA
  x
})

```

```{r na-quoated-recheck}

vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L))

```



On Zooplankton data


number of NAs per column
Map on multiple variables
transforming data in > column (e.g., [max / max-min])
using v-or-lapply to plot many Ys against one X variable
split-apply-do_something-combine

