---
title: ''
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Style Guide

PascalCase: function names; dataframes; field headings; file names  
camelCase: function parameters  
lower_case (with underscore): variables  

use `<-` instead of `=` for variable assignment  
most everytime function call, hard return after left parentheses  
each line keep to max 80 characters  

```
# create function
LengthFrequency <- function(data, speciesName, forkLength) {
  # do something here
}
# end LengthFrequency

len_freq <- LengthFrequency(
  data = Sturgeon,
  speciesName = "white",
  # speciesName = "green",
  forkLength = TRUE
)
```
# Functional Programming

- function call
- arguments can be other functions
- functions can return functions

# Function Call

`iris` built-in dataframe in `datasets` package

```{r func-call}

`+`(9, 8)                      # 9 + 8

`<-`(x, 1:10)                  # x <- 1:10 or x = 1:10
print(x)

`==`(9, 8)                     # 9 == 8

`[`(iris, 1, 2)                # iris[1, 2]
`[[`(iris, 1, 2)               # iris[[1, 2]]

`[[`(iris, 1, "Sepal.Width")   # iris[[1, "Sepal.Width"]]

# Not run --------------------------------------------------
# `[`(iris, 5)                 # iris[5]
# `[`(iris, , 5, drop = TRUE)  # iris[, 5, drop = TRUE]
# `[[`(iris, 5)                # iris[[5]]

```

# `for` Loop

```
for (items in variable) {
  # do something
}
```

```{r for-loop}
for (i in 1:5) {
  print(i)
}
```

# `<_>apply` Family Functions

```
lapply(list or vector, FUN = function(x) # do something)
```
# `lapply` Inner Workings

```{r lapply-show}

lapply

```

# `lapply` First Example

```{r fun-lapply}

lapply(1:5, FUN = `+`, 5)

```

# Vectorized

```{r vectorized}
1:5 + 5
```

# `lapply` On List

```{r list-create}

X <- list(
  Item1 = 1:5,
  Item2 = 6:10,
  Item3 = 11:20
)

X

```

because `X` is a list, vectorization concept doesn't work...

```{r list-add5, eval=FALSE}

X + 5
# Error in X + 5 : non-numeric argument to binary operator

```

so, here's where `lapply` is useful.

```{r list-lapply}

lapply(X, FUN = `+`, 5)
lapply(X, FUN = `^`, 2)

```

useful to get count for each item in list

```{r length-X}

lapply(X, FUN = length)

```

# `vapply` Inner Workings

```{r vapply-show}

vapply

```

# `vapply` Example

```{r vapply-example}

vapply(X, FUN = length, FUN.VALUE = numeric(1L))

```

now experiment with changing argument to `FUN` or setting `USE.NAMES = FALSE`

# More with `vapply`

```{r employees-list}

employees <- list(
  c("first", "last"),
  c("first", "last"),
  c("first", "last"),
  c("first", "last"),
  c("first")
)

employees

```

can we get employees' first name or concatenate `first` & `last`...

```{r employees-demo}

vapply(employees, FUN = `[`, FUN.VALUE = character(1L), 1)
vapply(employees, FUN = paste0, FUN.VALUE = character(1L), collapse = " ")

```

let's display our numeric list again

```{r X-display}

# display list `X` again for convenience
X

```

...what if we wanted some descriptive stats for each item in `X`

- how might you handle `NAs`
- how might you change output's orientation (i.e., stat names as columns)

```{r vapply-matrix_out1} 

vapply(X, FUN = function(x) {
  c(Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(3L))

```

same idea, but now we've included `N` or count

```{r vapply-matrix_out2}

vapply(X, FUN = function(x) {
  c(N = length(x), Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(4L))

```

probably the preferred orientation

```{r vapply-matrix_out3} 

t(vapply(X, FUN = function(x) {
  c(N = length(x), Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(4L)))

```

```{r vapply-list_out, eval=FALSE, echo=FALSE}

t(vapply(X, FUN = function(x) {
  list(N = length(x), Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = list(N = NA, Min = NA, Max = NA, Mean = NA)))

```

# List of Functions

```{r list-functions}

funs <- list(
  Mean = mean,
  Sd = sd,
  N = length,
  NumValsLT5 = function(x) sum(x < 5)
)

vapply(funs, FUN = function(f, nums) {
  f(nums)
}, FUN.VALUE = numeric(1L), nums = 1:10)

```

```{r data-download}

tf <- tempfile(pattern = "thisone", fileext = ".xlsx")

download.file(
  url = "ftp://ftp.wildlife.ca.gov/IEP_Zooplankton/1972-2018MysidMatrix.xlsx",
  # destfile = tempfile(fileext = ".xlsx")
  destfile = tf,
  mode = "ab"
)

Zoop <- readxl::read_excel(
  path = tf,
  sheet = "Mysid CPUE Matrix 1972-2018 "
)

unlink(tf)

```

```{r na-check}

vapply(Zoop, FUN = function(x) {
  sum(is.na(x))
}, FUN.VALUE = numeric(1L))

t(noquote(vapply(Zoop, FUN = function(x) {
  c(NumNA = sum(is.na(x)), TypeOf = typeof(x))
}, FUN.VALUE = character(2L))))

```

```{r summary-plus}

lapply(Zoop, FUN = function(x) {
  if (is.double(x)) return(range(x, na.rm = TRUE))
  table(x, useNA = "ifany", dnn = NULL)
})

```

```{r na-quoted}

vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L))

```

```{r char-nas}

char_nas <- vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L), USE.NAMES = FALSE)

b <- char_nas > 0

head(Zoop[b])

Zoop[b] <- lapply(Zoop[b], FUN = function(x) {
  x[x %in% "NA"] <- NA
  x
})

```

```{r na-quoated-recheck}

vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L))

```

```{r scaled}

cols <- c("N_ mercedis", "A_ hwanhaiensis")

fv <- nrow(Zoop)

apply(Zoop[cols], MARGIN = 2, FUN = range)

zoop_scaled <- vapply(Zoop[cols], FUN = function(x) {
  x / (max(x) - min(x))
}, FUN.VALUE = numeric(fv))

apply(zoop_scaled, MARGIN = 2, FUN = range)


```

```{r plot-not-scaled}

plot(x = Zoop[["Year"]], y = Zoop[[cols[1]]])
points(x = Zoop[["Year"]], y = Zoop[[cols[2]]], col = "red")

```

```{r plot-scaled}

plot(x = Zoop[["Year"]], y = zoop_scaled[, "N_ mercedis"])
points(x = Zoop[["Year"]], y = zoop_scaled[, "A_ hwanhaiensis"], col = "red")

```

```{r mean-temp-yr1}

mean_temp_yr <- with(data = Zoop, expr = {
  
  s <- split(Temperature, f = Year)
  
  vapply(s, FUN = mean, FUN.VALUE = numeric(1L))
  
})

mean_temp_yr

```

```{r mean-temp-yr2}

mean_temp_yr <- with(data = Zoop, expr = {
  
  s <- split(Temperature, f = Year)
  
  vapply(s, FUN = mean, FUN.VALUE = numeric(1L), na.rm = TRUE)
  
})

mean_temp_yr

```

```{r mean-temp-yr3}

mean_temp_yr <- with(data = Zoop, expr = {
  
  s <- split(Temperature, f = Year)
  
  # t(vapply(s, FUN = function(x, ...) {
  #   c(Mean = mean(x, ...), Sd = sd(x, ...))
  # }, FUN.VALUE = numeric(2L), na.rm = TRUE))
  
  t(vapply(s, FUN = function(x) {
    nna <- sum(is.na(x))
    x <- Filter(f = Negate(is.na), x = x)
    c(N = length(x), NNA = nna, Mean = mean(x), Var = var(x))
  }, FUN.VALUE = numeric(4L)))
  
})

str(mean_temp_yr)
tail(mean_temp_yr)

```

```{r mean-temp-yr-mon1}

mean_temp_yr_mon <- with(data = Zoop, expr = {
  
  # extract month from Date (as "01", "02", "03", etc.)
  m <- format(SampleDate, format = "%m")
  
  s <- split(Temperature, f = list(Year, m), drop = FALSE) #, sep = "_"
  
  # t(vapply(s, FUN = function(x, ...) {
  #   c(Mean = mean(x, ...), Sd = sd(x, ...))
  # }, FUN.VALUE = numeric(2L), na.rm = TRUE))
  
  t(vapply(s, FUN = function(x) {
    nna <- sum(is.na(x))
    x <- Filter(f = Negate(is.na), x = x)
    c(N = length(x), NNA = nna, Mean = mean(x), Var = var(x))
  }, FUN.VALUE = numeric(4L)))
  
})

dim(mean_temp_yr_mon)
dimnames(mean_temp_yr_mon)[[2]]
head(dimnames(mean_temp_yr_mon)[[1]], n = 10)

```

```{r mean-temp-yr-mon-show}

head(mean_temp_yr_mon)

```

```{r mean-temp-yr-mon2}

mean_temp_yr_mon <- with(data = Zoop, expr = {
  
  # extract month from Date (as "01", "02", "03", etc.)
  m <- as.numeric(format(SampleDate, format = "%m"))
  
  out <- aggregate(
    Temperature,
    by = list(Year = Year, Mon = m),
    FUN = function(x) {
      nna <- sum(is.na(x))
      x <- Filter(f = Negate(is.na), x = x)
      c(N = length(x), NNA = nna, Mean = mean(x), Var = var(x))
    }
  )
  
  out
})

str(mean_temp_yr_mon)

```

```{r mean-temp-yr-mon3}

mean_temp_yr_mon <- with(data = Zoop, expr = {
  
  # extract month from Date (as "01", "02", "03", etc.)
  m <- as.numeric(format(SampleDate, format = "%m"))
  
  out <- aggregate(
    Temperature,
    by = list(Year = Year, Mon = m),
    FUN = function(x) {
      nna <- sum(is.na(x))
      x <- Filter(f = Negate(is.na), x = x)
      c(N = length(x), NNA = nna, Mean = mean(x), Var = var(x))
    }
  )
  
  data.frame(
    out[c("Year", "Mon")],
    out[["x"]]
  )
})

str(mean_temp_yr_mon)

```

```{r plot-mult-y1}

b <- grepl(pattern = "^[A-Z]\\_", x = colnames(Zoop))
nms <- colnames(Zoop)[b]
nms

vapply(nms, FUN = function(y, d, xVar) {
  
  plot(x = d[[xVar]], y = d[[y]], xlab = xVar, ylab = y)
  
  mean(d[[y]], na.rm = TRUE)
  
}, FUN.VALUE = numeric(1L), d = Zoop, xVar = "Year")

```

```{r plot-mult-y2}

par(mfrow = c(4, 2))

vapply(nms, FUN = function(y, d, xVar) {
  
  plot(x = d[[xVar]], y = d[[y]], xlab = xVar, ylab = y)
  
  mean(d[[y]], na.rm = TRUE)
  
}, FUN.VALUE = numeric(1L), d = Zoop, xVar = "Secchi")

```

