---
title: ''
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Style Guide

Herein, I present some quick points about my style guide. This is by no means the best approach but one that has worked for me and one I've developed over the years.

PascalCase: function names; dataframes; field headings; file names  
camelCase: function parameters  
lower_case (with underscore): variables  

use `<-` instead of `=` for variable assignment  
most everytime function call, hard return after left parentheses  
each line keep to max 80 characters  

Below is an example creating a function, and then calling that function. Note how in the call each function parameter is on its own line. This allows for easily commenting a parameter to try another argument (see `speciesName`).

```
# create function
LengthFrequency <- function(data, speciesName, forkLength) {
  # do something here
}
# end LengthFrequency

len_freq <- LengthFrequency(
  data = Sturgeon,
  speciesName = "white",
  # speciesName = "green",
  forkLength = TRUE
)
```
# Functional Programming

`R` is a functional programming language. As such, it has certain characteristics that can work to one's advantage.

- everything is a function call
- arguments can be other functions
- functions can return functions

# Function Call

Here, we use `iris`, the built-in dataframe in `datasets` package. Below demonstrates the 'everything is a function call' aspect. Commented sections display the much more common use.

```{r func-call}

`+`(9, 8)                      # 9 + 8

`<-`(x, 1:10)                  # x <- 1:10 or x = 1:10
print(x)

`==`(9, 8)                     # 9 == 8

`[`(iris, 1, 2)                # iris[1, 2]
`[[`(iris, 1, 2)               # iris[[1, 2]]

`[[`(iris, 1, "Sepal.Width")   # iris[[1, "Sepal.Width"]]

# Not run --------------------------------------------------
# `[`(iris, 5)                 # iris[5]
# `[`(iris, , 5, drop = TRUE)  # iris[, 5, drop = TRUE]
# `[[`(iris, 5)                # iris[[5]]

```

To know the what-why-when's of the  `<l|v|m>apply()` family of `package:base` functions, it helps to understand --- at least conceptually --- about a `for` loop. For some items in variable `x`, do something.

# `for` Loop

```
for (items in variable) {
  # do something
}
```

...an example

```{r for-loop}
for (i in 1:5) {
  print(i)
}
```

# `<_>apply` Family Functions

Here in `lapply` the argument to `FUN` is that 'do something' in the `for` loop. The function passed to `FUN`, will act on every item in the `list or vector`. 

```
lapply(list or vector, FUN = function(x) # do something)
```
# `lapply` Inner Workings

```{r lapply-show}

lapply

```

# `lapply` First Example

Here we supply the vetor 1-->5. We wish to add 5 to each number in that vector.

```{r fun-lapply}

lapply(1:5, FUN = `+`, 5)

```

# Vectorized

But because R is vectorized, we can simply write below. The results remain neatly in a vector.

```{r vectorized}
1:5 + 5
```

# `lapply` On List

But what if instead of a vector, we had a **list** of vectors? (an uneven list in this case)

```{r list-create}

X <- list(
  Item1 = 1:5,
  Item2 = 6:10,
  Item3 = 11:20
)

X

```

because `X` is a list, vectorization concept doesn't work...

```{r list-add5, eval=FALSE}

X + 5
# Error in X + 5 : non-numeric argument to binary operator

```

so, here's where `lapply` is useful. (Adding `5` may not be of particular importance, but what about sqaring each value?)

```{r list-lapply}

lapply(X, FUN = `+`, 5)
lapply(X, FUN = `^`, 2)

```

We can certainly supply a built-in function to `FUN`. Sometimes it is useful to get count for each item in list.

```{r length-X}

lapply(X, FUN = length)

```

# `vapply` Inner Workings

`vapply` is quite similar to `lapply` but includes some extra (and convenient) parameters.

```{r vapply-show}

vapply

```

# `vapply` Example

We can tell `vapply` the expected output of `FUN`. We know `base::length()` returns a single numeric value, so we can supply just that to `FUN.VALUE`. *Note*: the `L` maintains 1 as an integer value, not a double.)

```{r vapply-example}

vapply(X, FUN = length, FUN.VALUE = numeric(1L))

```

Now experiment with changing argument to `FUN` or setting `USE.NAMES = FALSE`.

# More with `vapply`

Here we have a fake employee list. We don't have a last name for employee 5.

```{r employees-list}

employees <- list(
  c("first", "last"),
  c("first", "last"),
  c("first", "last"),
  c("first", "last"),
  c("first")
)

employees

```

Can we get employees' first name or concatenate `first` & `last`...again, remember that *everything* is a function call. Thus supplying `[` to `FUN` works. The `1` indicates the elment number we wish to index.

```{r employees-demo}

vapply(employees, FUN = `[`, FUN.VALUE = character(1L), 1)
vapply(employees, FUN = paste0, FUN.VALUE = character(1L), collapse = " ")

```

Let's display our numeric list again.

```{r X-display}

# display list `X` again for convenience
X

```

...what if we wanted some descriptive stats for each item in `X`

- how might you handle `NAs`?
- how might you change output's orientation (i.e., stat names as columns)?

```{r vapply-matrix_out1} 

vapply(X, FUN = function(x) {
  c(Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(3L))

```

same idea, but now we've included `N` or count

```{r vapply-matrix_out2}

vapply(X, FUN = function(x) {
  c(N = length(x), Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(4L))

```

probably the preferred orientation

```{r vapply-matrix_out3} 

t(vapply(X, FUN = function(x) {
  c(N = length(x), Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = numeric(4L)))

```

```{r vapply-list_out, eval=FALSE, echo=FALSE}

t(vapply(X, FUN = function(x) {
  list(N = length(x), Min = min(x), Max = max(x), Mean = mean(x))
}, FUN.VALUE = list(N = NA, Min = NA, Max = NA, Mean = NA)))

```

# List of Functions

We could pass a list of functions as an argument to `vapply`'s `X`. Though this convention is not as common. *Note*: the last function in the list is a custom function that returns count of values less than 5. `x < 5` returns a Boolean, and summing a Boolean returns number `TRUE`, as `TRUE` = 1, `FALSE` = 0.

```{r list-functions}

funs <- list(
  Mean = mean,
  Sd = sd,
  N = length,
  NumValsLT5 = function(x) sum(x < 5)
)

vapply(funs, FUN = function(f, nums) {
  f(nums)
}, FUN.VALUE = numeric(1L), nums = 1:10)

```

# Download Data

Here we use zooplankton data from CDFW's ftp site.

** note the use of package `readxl` ** (install prior to running below)

```{r data-download}

# tempdir()
# list.files(tempdir())

tf <- tempfile(pattern = "thisone", fileext = ".xlsx")

download.file(
  url = "ftp://ftp.wildlife.ca.gov/IEP_Zooplankton/1972-2018MysidMatrix.xlsx",
  # destfile = tempfile(fileext = ".xlsx")
  destfile = tf,
  mode = "ab"
)

# note the space after 2018 in the sheet name --- needed because that space
# exists in the .xlsx file
Zoop <- readxl::read_excel(
  path = tf,
  sheet = "Mysid CPUE Matrix 1972-2018 "
)

# always a good idea to remove (delete) the file after downloading data
unlink(tf)

```

# Check Data

The first thing we want to know about our data is...is it clean? Where are the NA values (in which fields)? What are field data types? Are these the correct data types? ...

`vapply` can help with this.

In this first example, we check each field in `Zoop` for NA values. The function will return 0 if no values are NA, and a value > 0 otherwise. The value > 0 indicates the number of NA records in that field.

```{r na-check1}

vapply(Zoop, FUN = function(x) {
  sum(is.na(x))
}, FUN.VALUE = numeric(1L))

```

Here, we include `typeof` to our output. `vapply` no returns a matrix. *Note*: we use `base::noquote` to remove quotes from the output (`typeof` returns a character), and `base::t` to transpose the matrix as two columns instead of two rows.

```{r na-check2}

t(noquote(vapply(Zoop, FUN = function(x) {
  c(NumNA = sum(is.na(x)), TypeOf = typeof(x))
}, FUN.VALUE = character(2L))))

```

# Custom Data Summary

Of course, using `utils::str` or `base::summary` on new data is always a good idea. Here, though, we can customize our summary, if you will. For each field in `Zoop`, if value is a double, return the range. Otherwise, return frequency table.

(**waring**: long output)

```{r summary-plus}

lapply(Zoop, FUN = function(x) {
  if (is.double(x)) return(range(x, na.rm = TRUE))
  table(x, useNA = "ifany", dnn = NULL)
})

```

We notice something interesting about fields `EZStation` and `DWRStation`. We have some quoted NAs. That is, entries in the Excel file are `NA`, not just blank (which would be read by R as `r NA`). This certainly poses a problem for data summary and other analytics.

```{r na-quoted}

vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L))

```

# Clean Data

So how do we efficiently replace the quoted (or character) NA with R's `r NA`?

(1) `char_nas` returns the number of character NAs in each field
(2) `b` sets to `TRUE` any value > 0; `head(Zoop[b])` displays n=6 for quick viewing
(3) change all character NAs to `r NA` using `lapply`

*Note*: using `Zoop[b]` before assignment `<-` maintains the data.frame structure. Otherwise, you will get a list.

*Note*: instead of anonymous function, below would be much better suited as a named function. This would allow for ease of use on other data.frames

```
function(x) {
  x[x %in% "NA"] <- NA
  x
}
```

```{r char-nas}

char_nas <- vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L), USE.NAMES = FALSE)

b <- char_nas > 0

head(Zoop[b])

Zoop[b] <- lapply(Zoop[b], FUN = function(x) {
  x[x %in% "NA"] <- NA
  x
})

```

Now let us do a quick recheck. All character `NAs` should be gone, yes?

```{r na-quoted-recheck}

vapply(Zoop, FUN = function(x) {
  sum(x %in% "NA")
}, FUN.VALUE = numeric(1L))

```

# Scale Data

Sometimes we wish to compare variables with markedly different ranges. Here fields `N_ mercedis` and `A_ hwanhaiensi` fit that bill.

```{r scaled1}

cols <- c("N_ mercedis", "A_ hwanhaiensis")

vapply(Zoop[cols], FUN = range, FUN.VALUE = numeric(2L))

# works too
# apply(Zoop[cols], MARGIN = 2, FUN = range)

```

Here we standardize ([min-max scaling](https://en.wikipedia.org/wiki/Feature_scaling)) the desired fields dividing each value (minus the minimum value) by the range difference.

$$\frac{x-min(x)}{max(x)-min(x)}$$

```{r scaled2}

# for vapply output
fv <- nrow(Zoop)

zoop_scaled <- vapply(Zoop[cols], FUN = function(x) {
  (x - min(x)) / (max(x) - min(x))
}, FUN.VALUE = numeric(fv))

# verify range is 0 to 1
apply(zoop_scaled, MARGIN = 2, FUN = range)

```

Plotting the raw data compared to the standardized, we get the following.

```{r plot-not-scaled}

plot(x = Zoop[["Year"]], y = Zoop[[cols[1]]])
points(x = Zoop[["Year"]], y = Zoop[[cols[2]]], col = "red")

```

```{r plot-scaled}

plot(x = Zoop[["Year"]], y = zoop_scaled[, "N_ mercedis"])
points(x = Zoop[["Year"]], y = zoop_scaled[, "A_ hwanhaiensis"], col = "red")

```

*Note*: standardizing and (or) plotting these data in this manner might not be visually or statistically appropriate. This is merely for demonstration purposes.

# Data Split

Sometimes we wish to summarize data by a particular variable or variable grouping. `stats::aggregate` is always a nice option. Here, though, we demonstrate `base::split`, which returns a list and thus works nicely with `<l|v>apply`.

#### Example 1: mean water temperature by year

Here we use `base::with` for convenience (i.e., to limit `Zoop$Temperature` or `Zoop[["Temperature"]]`). Variable `s` is a list of temperature values by year. `vapply` takes each element in the list (i.e., each year) and returns a mean value.

```{r mean-temp-yr1}

mean_temp_yr <- with(data = Zoop, expr = {
  
  s <- split(Temperature, f = Year)
  
  vapply(s, FUN = mean, FUN.VALUE = numeric(1L))
  
})

mean_temp_yr

```

Note, however, in the output above, we have many `NA`s. We can use the `...` in `vapply` to pass `na.rm = TRUE` to `base::mean`. Better...

```{r mean-temp-yr2}

mean_temp_yr <- with(data = Zoop, expr = {
  
  s <- split(Temperature, f = Year)
  
  vapply(s, FUN = mean, FUN.VALUE = numeric(1L), na.rm = TRUE)
  
})

mean_temp_yr

```

Suppose, however, we wanted more than just mean. Typically, along with mean we would like some measure of spread (e.g., variance) and some count (e.g., n).

We can still use `vapply`, but now we've replaced `base::mean` with an anonymous function that includes `mean`. (*Note*: here is where a custom function comes in very handy.)

The anonymous function is this...

```
function(x) {
  nna <- sum(is.na(x))
  x <- Filter(f = Negate(is.na), x = x)
  c(N = length(x), NNA = nna, Mean = mean(x), Var = var(x))
}
```

Let us describe each line.

`nna <- sum(is.na(x))` assigns to variable `nna` the count of values that are `NA`

`x <- Filter(f = Negate(is.na), x = x)` reassigns variable `x` but without `NA`s (for help run `?Filter` in your R console).

- `N = length(x)` gives count of values in `x`, less `NA`s

- `NNA = nna` gives count of values in `x`, **including** `NA`s (remember, we removed `NA`s *after* `sum(is.na(x))`)

- `Mean = mean(x)` returns the mean value of `x` (`na.rm = TRUE` is not needed)

- `Var = var(x)` returns the variance value of `x` (`na.rm = TRUE` is not needed)

Finally, using `c()` bundles the output neatly into a numeric vector, length = 4, thus `FUN.VALUE = numeric(4L)` below. Optionally, a call to `t` manipulates output such that the *rows* represent year and *columns* `N`, `NNA`, `Mean`, and `Var`.

```{r mean-temp-yr3}

mean_temp_yr <- with(data = Zoop, expr = {
  
  s <- split(Temperature, f = Year)
  
  # t(vapply(s, FUN = function(x, ...) {
  #   c(Mean = mean(x, ...), Sd = sd(x, ...))
  # }, FUN.VALUE = numeric(2L), na.rm = TRUE))
  
  t(vapply(s, FUN = function(x) {
    nna <- sum(is.na(x))
    x <- Filter(f = Negate(is.na), x = x)
    c(N = length(x), NNA = nna, Mean = mean(x), Var = var(x))
  }, FUN.VALUE = numeric(4L)))
  
})

# just for display
str(mean_temp_yr)
tail(mean_temp_yr)
dim(mean_temp_yr)

```

#### Example 2: mean water temperature by year by month

Now suppose we wanted our output to be by year and month. Same thing basically, except we pass to `base::split` year & month (extracted from `SampleDate` using `base::format`). Note, however, the parameter `drop` to which we pass `FALSE`. This will include levels that did not occur. For example, no (sampling) data exists for January 1973, yet `mean_temp_yr_mon` will contain `0, 0, NaN, NA` as output for that year-month combination.

Experiment with setting `drop = TRUE`. What changes?

```{r mean-temp-yr-mon1}

mean_temp_yr_mon <- with(data = Zoop, expr = {
  
  # extract month from Date (as "01", "02", "03", etc.)
  m <- format(SampleDate, format = "%m")
  
  s <- split(Temperature, f = list(Year, m), drop = FALSE) #, sep = "_"
  
  # could be use but gets a little cumbersome
  # t(vapply(s, FUN = function(x, ...) {
  #   c(Mean = mean(x, ...), Sd = sd(x, ...))
  # }, FUN.VALUE = numeric(2L), na.rm = TRUE))
  
  t(vapply(s, FUN = function(x) {
    nna <- sum(is.na(x))
    x <- Filter(f = Negate(is.na), x = x)
    c(N = length(x), NNA = nna, Mean = mean(x), Var = var(x))
  }, FUN.VALUE = numeric(4L)))
  
})

# for display only
dim(mean_temp_yr_mon)
dimnames(mean_temp_yr_mon)[[2]]
head(dimnames(mean_temp_yr_mon)[[1]], n = 10)

```

```{r mean-temp-yr-mon-show}

head(mean_temp_yr_mon)

```

#### Example 3: employing `stats::aggregate`

Here we use the same anonymous function. But now we are using `aggregate`, which does the `split`ting for us. Again, we want water temperature mean and variance by year and month. For convenient output, we set month `as.numeric`.

We call `utils::str` and notice something interesting about the output. The structure is a `data.frame`, but our function return --- assigned to field `x` --- is a matrix. This is not really a problem, but if we want a "clean" data.frame, we can simply run the following code (see too output in the next chunk).

```
data.frame(
  out[c("Year", "Mon")],
  out[["x"]]
)
```

```{r mean-temp-yr-mon2}

mean_temp_yr_mon <- with(data = Zoop, expr = {
  
  # extract month from Date (as "01", "02", "03", etc.)
  m <- as.numeric(format(SampleDate, format = "%m"))
  
  out <- aggregate(
    Temperature,
    by = list(Year = Year, Mon = m),
    FUN = function(x) {
      nna <- sum(is.na(x))
      x <- Filter(f = Negate(is.na), x = x)
      c(N = length(x), NNA = nna, Mean = mean(x), Var = var(x))
    }
  )
  
  out
})

str(mean_temp_yr_mon)

```

```{r mean-temp-yr-mon3}

mean_temp_yr_mon <- with(data = Zoop, expr = {
  
  # extract month from Date (as "01", "02", "03", etc.)
  m <- as.numeric(format(SampleDate, format = "%m"))
  
  out <- aggregate(
    Temperature,
    by = list(Year = Year, Mon = m),
    FUN = function(x) {
      nna <- sum(is.na(x))
      x <- Filter(f = Negate(is.na), x = x)
      c(N = length(x), NNA = nna, Mean = mean(x), Var = var(x))
    }
  )
  
  data.frame(
    out[c("Year", "Mon")],
    out[["x"]]
  )
})

str(mean_temp_yr_mon)

```

# Plotting

We can use `vapply` in `plot`ting to loop through select variables. Say we wanted an annual time series of species CPUE (or whatever the data are).

We can "cheat" a bit to get species field names from our data. Each species --- less `Unidentified mysid` --- begins with a capital letter followed by an underscore. Using the regular expression `^[A-Z]\\_` and a call to `base::grepl`, we can select only those names (see `nms` variable below).

`nms` becomes the variable on which we will loop and will be the `y` variable in our time series plot. Other function parameters are `d` (our dataframe) and `xVar` (i.e., the `x` variable in our plot, to which in this case we set as `"Year"`).

LEFT OFF HERE 

```{r plot-mult-y1}

b <- grepl(pattern = "^[A-Z]\\_", x = colnames(Zoop))
nms <- colnames(Zoop)[b]
nms

vapply(nms, FUN = function(y, d, xVar) {
  
  plot(x = d[[xVar]], y = d[[y]], xlab = xVar, ylab = y)
  
  mean(d[[y]], na.rm = TRUE)
  
}, FUN.VALUE = numeric(1L), d = Zoop, xVar = "Year")

```

```{r plot-mult-y2, eval=FALSE}

par(mfrow = c(4, 2))

vapply(nms, FUN = function(y, d, xVar) {
  
  plot(x = d[[xVar]], y = d[[y]], xlab = xVar, ylab = y)
  
  mean(d[[y]], na.rm = TRUE)
  
}, FUN.VALUE = numeric(1L), d = Zoop, xVar = "Secchi")

```

