---
title: "IEP Data Manipulation and Processing"
author: "Rosemary Hartman"
date: "12/2/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

1. Importing data
1. Foundational skills of data manipulations
2. Detecting errors and outliers
3. Dataset restructuring
4. Missing data
5. Miscellaneous data manipulation
6. getting data out of R


Do we want to go into data organization at all? Maybe introduce "[tidy data](https://learning.nceas.ucsb.edu/2019-11-RRCourse/data-modeling-tidy-data.html)"?


## Importing data

There are multiple different ways of getting your data into R. RStudio 
has a nice little point=and-click interface that lets you import your dataset, however using
the GUI makes it more difficult to replicate your work later. Therefore, it's better
to use the comand-line to import your data.

We are going to use the Zooplankton Survey data for this demonstration. It's available 
[here](ftp://ftp.wildlife.ca.gov/IEP_Zooplankton/1972-2018CBMatrix.xlsx)

If you have a .csv file, you can use base R. So I went into the excel file and saved
one of the worksheets as a .csv

```{r imports}
?read.csv

CBdata = read.csv("CBdata.csv", stringsAsFactors = F)
head(CBdata)

```

But what if we want to import the origional excel file? For that,
we need the "readxl" package

```{r excel}

library(readxl)

?read_excel

CBdata = read_xlsx("1972-2018CBMatrix.xlsx", sheet = "CB CPUE Matrix 1972-2018")
#Yuck. If you have too many blanks in the begining of your sheet, it thinks it is
#a logical vector. Let's fix that.

CBdata = read_excel("1972-2018CBMatrix.xlsx", 
    sheet = "CB CPUE Matrix 1972-2018", col_types = c("numeric", 
        "numeric", "numeric", "numeric", 
        "date", "text", "text", "text", "numeric", 
        "text", "numeric", "text", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric", "numeric", "numeric", 
        "numeric"))

head(CBdata)

```
## Foundational skills

Do we really want to do data generation? Seems like a lot of work. 
###	Generating data
1.	rnorm, rbinom

### Dealing with data frames

A lot of these skills can be done with either Base R syntax, or "tidyverse"
syntax. Most people find "tidyverse" a little more intuitive, but you
can use whichever you are more comfortable with.

#### Renaming columns

```{r rename}
library(tidyverse)

names(CBdata)

#several options for how to rename your colums

names(CBdata)[names(CBdata)== "SurveyCode"] = "cheese"

names(CBdata)

#you could also just make a new column and delete the old one
CBdata$crackers = CBdata$cheese
CBdata$cheese = NULL

names(CBdata)

#using "rename" is a little more intuitive
CBdata = rename(CBdata, SurveyCode = crackers)

names(CBdata)

```

####	Reordering columns

But now our "SurveyCode" column is at the end. We want to put it back in the begining.

```{r}
#here's how to do it in base R
CBdata = CBdata[,c(74, 1:73)]

names(CBdata)

#you can also use the "select" function in dplyr
CBdata = select(CBdata, Survey, Year, SurveyCode, SurveyRep:CrabZoea)

names(CBdata)

#I just learned this cool thing when you have a lot of columns!
CBdatatest = select(CBdata, DAPHNIA, everything())

names(CBdatatest)

```


####	Adding new mutated data

We frequently need to transform data (especially log transformations). We 
might also need to calculate CPUE, or do other stuff, and add our calculated variables
to our data set. 

```{r mutate}

#we can just define a new variable and assign its value with the old variable
CBdata$ln_cals = log(CBdata$ALLCALADULTS + 1)

#or we can use the "mutate" function from dplyr. This is especially
#helpful if you want to make a bunch of variables at once

CBdata = mutate(CBdata, ln_clads = log(ALLCLADOCERA + 1), ln_cyc = log(ALLCYCADULTS +1), ln_all = ln_cals + ln_cyc + ln_clads)

View(CBdata)

```

#### converting among data classes

Even though we read in all our columns based on particular classses, sometimes
we want to change the data class. Sometimes that's super easy, sometimes it's more
difficult

```{r}

str(CBdata)

#changing things from characters to factors and back again is really easy


```

ix.	Converting among data classes
viii.	reordering factors


ii.	Apply functions

R goes fastest when you can apply a function to every element in
a vector (or a data frame)

```{r apply}


```
iii.	For loops
iv.	Aggregate()

x.	Convert numeric categorical data into text
1.	E.g., from 0/1 to “absent”/”present”
xi.	Working with Dates
1.	Converting to usable date notation
a.	The format() function
2.	How to generate DOY


```{r foundations}


```


