---
title: "Dates"
author: "Rosemary Hartman"
date: "06/08/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = F)
```
# Dates

Dealing with dates is one of the most difficult data types, but fortunately the package "lubridate" makes it much easier.

There are a few different types of data that refer to dates. "Dates" just have a date component, whereas date-times may be POSIXct (number of seconds since 1970), or PSIXlt (a named vector of year/month/day/hour/minute/second/time zone).



```{r}
library(tidyverse)
library(lubridate)
?lubridate

?DateTimeClasses

today()

now()

class(today())
class(now())

```
To play with dates, let's read in the CB data again, but we won't tell it that the "date" column is a date

```{r}
library(readxl)

CBdates = read.csv("CBdata.csv")[,1:10]
str(CBdates$Date)

```
It things "Date" is a vector of characters. We want to tell it that it's a date.

In base R, we can use as.date()

```{r}

zoopdates = as.Date(CBdates$Date)

str(zoopdates)
#Oops

?as.Date
?strptime


#if we don't tell it the format we are in trouble.
zoopdates = as.Date(CBdates$Date, format = "%m/%d/%Y")

str(zoopdates)

```
Well, that seemed to work, but I can never remember how the "format" thing works. Fortunately, lubridate has some shortcuts. 

```{r}
?ymd

CBdates = mutate(CBdates, Date2 = mdy(Date))
str(CBdates)

```
If I want to organize my dataset by month, I need to extract the "month" part of the date.

```{r}

?month

month(today())

month(today(), label = TRUE)

CBdates = mutate(CBdates, Month = month(Date2), 
                 julianday = yday(Date2), 
                 monthday = mday(Date2), 
                 weekday = wday(Date2))
View(CBdates)
```

# Now you try it!

Take 20-30 minutes to work through these examples. Feel free to raise your hand or use the chat to ask for help. Use the "thumbs up" reaction when you are done. 


1. What happens if you parse a string that contains invalid dates?

```{r}

ymd(c("2010-10-10", "cheese"))

```
2. Parse each of the following dates (use lubridate or base r, as you see fit)

```{r}
d1 <- "January 1, 2010"

d2 <- "2015-Mar-07"

d3 <- "06-Jun-2017"

d4 <- c("August 19 (2015)", "July 1 (2015)")

d5 <- "12/30/14" # Dec 30, 2014

```
3. We've dealt with the "date" column of the zooplankton dataset, what about the "time" column? Can you convert it to a date-time format?

```{r}

```

4. How do you add the date and time together into a single column?

```{r}

```
5. How often does the zooplankton survey sample on the weekends?

